Supplementary Methods for\\ \emph{Effectiveness of UNAIDS targets and HIV vaccination across 127 countries}

### 2016-10-19 Documents/supplementary_information.tex
Minor changes for Nature.
More changes for Nature.

\usepackage[font={sf}, labelfont={bf},
  labelsep=period, singlelinecheck=off]{caption}
\usepackage[font={sf}]{floatrow}
\usepackage{tabularx}

\usepackage[noblocks]{authblk}
\renewcommand{\Authsep}{, }
\renewcommand{\Authand}{ \& }
\renewcommand{\Authands}{ \& }
\renewcommand{\Affilfont}{\small}

@@

\usepackage[style=science, article-title=true, maxnames=100]{biblatex}
\addbibresource{supplementary_text.bib}
\newcommand{\labeltitlepunc}{}
\DeclareBibliographyDriver{incollection}{%
  \usebibmacro{bibindex}%
  \usebibmacro{begentry}%
  \usebibmacro{author/translator+others}%
  \setunit{\labelnamepunct}\newblock
  \printlist{language}%
  \newunit\newblock
  \usebibmacro{byauthor}%
  \newunit\newblock
  \usebibmacro{title}%
  \usebibmacro{in:}%
  \usebibmacro{maintitle+booktitle}%
  \newunit\newblock
  \usebibmacro{byeditor+others}%
  \newunit\newblock
  \printfield{volumes}%
  \newunit\newblock
  \printfield{note}%
  \newunit\newblock
  \usebibmacro{publisher+location+date}%
  \newunit\newblock
  \printfield{volume}
  \newunit
  \usebibmacro{chapter+pages}%
  \newunit\newblock
  \iftoggle{bbx:isbn}
    {\printfield{isbn}}
    {}%
  \newunit\newblock
  \usebibmacro{doi+eprint+url}%
  \newunit\newblock
  \usebibmacro{addendum+pubstate}%
  \setunit{\bibpagerefpunct}\newblock
  \usebibmacro{pageref}%
  \newunit\newblock
  \usebibmacro{related}%
  \usebibmacro{finentry}%
}

@@

\newcommand{\labelPrefix}{S}
\renewcommand{\thesection}{\labelPrefix\arabic{section}}
\renewcommand{\theequation}{\labelPrefix\arabic{equation}}
\renewcommand{\thefigure}{\labelPrefix\arabic{figure}}
\renewcommand{\thetable}{\labelPrefix\arabic{table}}
\renewcommand{\figurename}{Fig.}

\title{Supporting online material for\\
  \emph{Effectiveness of UNAIDS targets and HIV vaccination across 127
    countries}}

\author[1*]{Jan Medlock}
\author[2]{Abhishek Pandey}
\author[2]{Alyssa S.~Parpia}
\author[2]{Amber Tang}
\author[2]{Laura A.~Skrip}
\author[2]{Alison P.~Galvani}
\affil[1]{Department of Biomedical Sciences, Oregon State University,
  106 Dryden Hall, Corvallis, OR, 97331-4801, USA}
\affil[2]{Center for Infectious Disease Modeling and Analysis, Yale
  School of Public Health, 135 College Street, New Haven, USA}
\affil[*]{To whom correspondence should be addressed.  E-mail:
  \href{mailto:jan.medlock@oregonstate.edu}{
    \texttt{jan.medlock@oregonstate.edu}}}

@@

The model simulation and analysis tools, written in Python, are
publicly available at\\
\url{https://github.com/janmedlock/HIV-95-vaccine/}
\cite{medlock2016-git}.

@@

We developed a country-level continuous-time, compartmental model that
divides people aged $15$\;y and older into non-overlapping HIV states.
We chose to consider only these ages because they are the vast
majority of global PLHIV, 93\% in 2014 \cite{UNICEF}; because
population-level HIV estimates and demographic data are often
stratified into ages $15$--$49$; and to capture the continuing burden
of infections in people ages $50$ and older.

The background mortality rate ($\mu$) in each of the
127 countries was collected from published estimates
\cite{World_Development_Indicators2013-ee}.  The recruitment rate
($\kappa$) into the age $15\;\text{y}$ and older group was set by the
difference between the country's population growth rate
\cite{WorldBankpg} and its background mortality rate
\cite{World_Development_Indicators2013-ee}, as would be true if the
population was at its stable age structure.

Our model has 8 HIV states (\autoref{model_diag}): susceptible to HIV
infection ($S$ is the number of people susceptible), vaccinated
against HIV ($R$), acute HIV infection ($A$), undiagnosed HIV
infection ($U$), diagnosed but untreated HIV infection ($D$), treated
without viral suppression ($T$), viral suppression ($V$), and having
AIDS ($W$).  Viral suppression is defined as a viral load of less than
1000 viral RNA copies per mL of blood.  Viral suppression not only
extends survival but also dramatically reduces transmission
\cite{vernazza2000, ioannidis2001, attia2009, May2014-gp}.
Transitions between model states are governed by a series of
differential equations parametrized using values found in other
studies, along with transmission rates derived from UNAIDS estimates
of incidence and prevalence (\autoref{model_param}).  The model is
deterministic, but uncertainty in the parameters is treated by running
the model 1000 times with samples from the parameter distributions
(\autoref{uncertainty}) and summarizing the model outcomes using the
median, 1st and 3rd quartiles (i.e., 25th and 75th percentiles), and
5th and 95th percentiles.  Each country was modeled separately
(\hyperref[effectiveness_Afghanistan]{Figures
  \ref*{effectiveness_Afghanistan}}--\ref{effectiveness_Zimbabwe}) and
the country simulations were aggregated to calculate global and
regional outcomes
(\hyperref[effectiveness_Global]{Figures
  \ref*{effectiveness_Global}}--\ref{effectiveness_Western_Europe}).

@@

\begin{table}
  \begin{center}
    \begin{tabularx}{\textwidth}{lXlll}
      \hline
      & Definition & Value & Reference \\
      \hline
      $\mu$ & Background death rate & Country specific, Table S2
      & \cite{World_Development_Indicators2013-ee} \\
      $\kappa$ & Recruitment rate & Country specific, Table S2
      & \cite{World_Development_Indicators2013-ee, WorldBankpg} \\
      $\delta$	& Rate of leaving acute infection
      & $\Triangular$(2, 4.14, 9.6)\;y$^{-1}$
      & \cite{Hollingsworth2008-iy} \\
      $\sigma$	& Rate of developing AIDS without viral suppression
      & 0.1064\;y$^{-1}$ & \cite{Morgan2002-cq} \\
      $\theta$ & Rate of developing AIDS with viral suppression
      & Eq.~(\ref{theta}) & --- \\
      $\gamma$ & Rate of viral suppression
      & $\Uniform$(0.5, 1.5)\;y$^{-1}$
      & \cite{Currie2009-yz} \\
      $\nu$ & Death rate from AIDS & 0.5\;y$^{-1}$
      & \cite{Morgan2002-cq} \\
      $\rho$ & Vaccine efficacy & 50\%\,[30\%, 70\%] & --- \\
      $\omega$	& Reduction in lifetime with viral suppression
      & $\Uniform$(5, 8)\;y
      & \cite{Samji2013-kf, Unaids2014-ue} \\
      $\tau_{A}$ & Transmissibility during acute phase
      & $\Triangular$(0.0039, 0.0082, 0.0150)
      & \cite{Wawer2005-us, Skarbinski2015-ni}\\
      $\tau_{U}$ & Transmissibility after acute phase
      & $\Triangular$(0.00077, 0.0014, 0.00251)
      & \cite{Hughes2012-so} \\
      $\varepsilon$ & Relative transmissibility with
      viral suppression & $\BetaPERT$(0.08, 0.002, 0.57)
      & \cite{Donnell2010-xo, attia_2009, wilson_2012, jia_2013,
        rodger_2016} \\
      $n$ & Coital acts per year & $\Uniform$(96, 108)
      & \cite{Wawer2005-us, Abdool_Karim2010-cm}\\
      $\alpha$ & Diagnosis rate & See eq.~(\ref{diagnosis_rate}) & --- \\
      $\phi$ & Treatment rate & See eq.~(\ref{treatment_rate}) & --- \\
      $\psi$ & Rate of relapse to untreated & See eq.~(\ref{relapse_rate})
      & --- \\
      $\zeta$ & Vaccination rate & See eq.~(\ref{vaccination_rate}) & --- \\
      \hline
    \end{tabularx}
    \caption{Model parameters. $\Triangular$, $\Uniform$, and
      $\BetaPERT$ are sampling distributions: see
      \autoref{uncertainty} for a full description.  For vaccine
      efficacy, $50\%$ was the baseline, and $30\%$ and $70\%$ were
      also used in a scenario analysis.}
    \label{model_param}
  \end{center}
\end{table}

People susceptible to HIV ($S$) progress to acute infection ($A$)
according to the force of infection ($\lambda$,
eq.~\eqref{force_of_infection}), which depends on the HIV transmission
rate, estimated from the country-specific incidence and prevalence
data (\autoref{model_fitting}), adjusted for the current number of
acutely infected, untreated or ineffectively treated, and viral
suppression PLHIV.  Acute infection is characterized by high viral
loads and lasts an average of 2.90 [95\% confidence interval 1.23,
6.00]\;months \cite{Hollingsworth2008-iy}.  (We converted these
durations to rates like $1 / 2.90\;\text{month$^{-1}$} \approx
4.14\;\text{y$^{-1}$}$ and set the parameter using random samples from
$\Triangular(2, 4.14, 9.6)\;\text{y$^{-1}$}$ to capture the
uncertainty in its value.  See \autoref{uncertainty} for the
definition of $\Triangular(a, b, c)$ and other distributions.)  We
assumed that people remain undiagnosed during their acute-infection
phase. After acute infection, people move into the chronic undiagnosed
class ($U$).  People in the undiagnosed class will be diagnosed at the
rate $\alpha$, which is determined by the country's current diagnosis
level and its diagnosis target (\autoref{targets}).  Diagnosed people
($D$) transition to the treated compartment ($T$) upon starting ART at
the rate $\phi$, dependent on the country's current treatment level
and its treatment target (\autoref{targets}).  After some time,
treated people then transition to the viral-suppression class ($V$),
at the rate $\gamma$.  We took $\gamma$ as a random sample from
$\Uniform(0.5, 1.5)\;\text{y$^{-1}$}$, that is, viral suppression
occurs after between 8 months and 2 years on treatment
\cite{Currie2009-yz}.  Disengagement from treatment moves people from
treated ($T$) and viral suppression ($V$) at the rate $\psi$,
dependent on the current level of viral suppression and the country's
target (\autoref{targets}), back to the diagnosed class ($D$).

PLHIV without viral suppression (compartments $U$, $D$, and $T$)
develop AIDS (compartment $W$) after an average duration of 10.4 y
\cite{Morgan2002-cq}, i.e., with transition rate
$\sigma = 0.1064\;\text{y$^{-1}$}$.  People who have achieved viral
suppression ($V$) develop AIDS at rate $\theta$, slower than PLHIV
without viral suppression.  People with viral suppression are
estimated to live $5$--$8$\;y shorter lives than the non-HIV infected
\cite{Samji2013-kf, Unaids2014-ue}, which we quantified by randomly
sampling the parameter $\omega$ from $\Uniform(5, 8)\;\text{y}$.
Assuming no relapse to untreated, the duration of viral suppression is
\begin{equation}
  \frac{1}{\theta + \mu} = \frac{1}{\mu} - \omega - \frac{1}{\nu}.
\end{equation}
The term to the left of the equals sign is the time until leaving
viral suppression by either progression to AIDS (at rate $\theta$) or
background mortality (at rate $\mu$).  The term to the right of the
equals sign is the lifespan of the non-HIV infected, minus the
reduction in life due to having HIV with viral suppression, minus the
duration of the AIDS stage.  Solving for $\theta$ gives the rate of
progression to AIDS with viral suppression as
\begin{equation}
  \label{theta}
  \theta = \frac{1}{1/\mu - \omega - 1/\nu} - \mu.
\end{equation}

@@

The relative transmission rates of acute, unsuppressed, and suppressed
infected people are
\begin{equation}
  \label{betas}
  \begin{split}
    \beta_A &= 1 - (1 - \tau_A)^n,
    \\
    \beta_U &= 1 - (1 - \tau_U)^n,
    \\
    \beta_V &= 1 - (1 - \varepsilon \tau_U)^n,
  \end{split}
\end{equation}
where the $\tau$'s are the relative transmissibilities of people with
acute ($A$), unsuppressed ($U$ and $T$), and suppressed ($V$)
infections, $n$ is the annual number of coital acts, and $\varepsilon$
is the relative transmissibility with viral suppression.  For
transmissibility during the acute phase, we sampled $\tau_A$ from
$\Triangular$(0.0039, 0.0082, 0.0150)
\cite{Wawer2005-us, Skarbinski2015-ni}, whereas, for transmissibility
for unsuppressed people, we sampled $\tau_U$ from
$\Triangular(0.00077, 0.0014, 0.00251)$ \cite{Hughes2012-so}.  The
transmissibility of people with viral suppression relative to those
without suppression has been estimated at 0.08 (95\% confidence
interval 0.002, 0.57) \cite{Donnell2010-xo}: due to the extremely long
left tail, we chose to sample the relative transmissibility
$\varepsilon$ from $\BetaPERT(0.08, 0.002, 0.57)$ rather than a
$\Triangular$ distribution.  (See \autoref{uncertainty} for a full
description of $\BetaPERT(a, b, c)$.)  The annual number of coital
acts $n$ was sampled from $\Uniform(96, 108)$
\cite{Wawer2005-us, Abdool_Karim2010-cm}.  The country-specific
transmission rate $\eta$ was derived from UNAIDS estimates of each
country's longitudinal HIV prevalence and incidence
(\autoref{model_fitting}).

@@

The parametrized differential equations were solved numerically
\cite[see][]{medlock2016-git} and four effectiveness outcomes were
computed from the solutions.
\begin{description}
\item[New infections] New infections from 2015 to time $t$ are given
  by $Y(t)$.

@@

The results from the country simulations were aggregated to the
regional and global levels by adding the numbers of people in each
compartment over time and then new infections, per-capita incidence,
PLHIV, and HIV-related deaths were calculated from the aggregates.
(See Table S6 for the definitions of UNAIDS regions.)

@@

We modeled 3 different targets for the proportion diagnosed ($p_D$), the
proportion treated ($p_T$), and the proportion with viral suppression ($p_V$).
\begin{description}
\item[Status quo] The status quo target is for the proportions $p_D$,
  $p_T$, and $p_V$ to remain fixed at their initial levels (Table S5)
  going forward. That is,
  \begin{equation}
    \label{status_quo_target}
    \begin{split}
      p_D^*(t) &= p_D(2015), \\
      p_T^*(t) &= p_T(2015), \\
      p_V^*(t) &= p_V(2015).
    \end{split}
  \end{equation}

@@

For 127 countries, we found sufficient published estimates and data to
parametrize our model.  Demographic data, including population growth
rate \cite{WorldBankpg}, death rate
\cite{World_Development_Indicators2013-ee}, and number of people aged
15--49 years \cite{The_World_Bank2016-fd} were obtained from the World
Bank (Table S2). Longitudinal HIV prevalence (Table S3) and incidence
(Table S4) estimates for ages 15--49, spanning from as early as 1990
to 2015, were primarily obtained from the AIDSinfo database produced
by UNAIDS \cite{Unaids2016-an}. Other sources included UNAIDS Country
Progress Reports \cite{Unaids2016-am} published from 2012 to 2016 and
AIDS Data Hub \cite{AIDSdatahub-fg}, which compiles data from UNAIDS,
UNICEF, WHO, and the Asian Development Bank. These sources were also
used to inform the initial conditions for the model: the number of
people diagnosed with HIV, the number on ART, and the number who have
viral suppression or have been retained on treatment for at least 12
months (Table S5). Where estimates or data were not available from
these sources, alternative resources such as peer-reviewed journal
articles and country health ministry reports were consulted.  See
Tables S2--S5 for full information on the sources used for each
country.  These source tables are also available in the public
source-code repository \cite{medlock2016-git}.

@@

Using available historical country-level estimates of prevalence and
incidence (Tables S3 and S4), and published estimates of
transmissibility
\cite{Wawer2005-us,Donnell2010-xo,Hughes2012-so,Skarbinski2015-ni}, we
derived an average country-specific transmission rate, which then
informs the acute, unsuppressed, and suppressed transmission rates.
This calibration method provided a recent estimate of country-specific
transmission, while using historical data to smooth out short-term
variations.

@@

To reflect the uncertainty in $\beta(t)$, we assumed that it follows a
lognormal distribution (\autoref{uncertainty}). For the parameters
$\mu$ and $\sigma^2$ of the lognormal distribution we used the
exponentially weighted mean and variance \cite{holt2004} over time
(with half-life 1\;y) of $\log \beta(t)$ for the year 2015.  That is,
\begin{equation}
  \label{lognormal_params}
  \begin{split}
    \mu &= \frac{\sum_{i = 1}^n  2^{t_i}
      \log \beta(t_i)}
    {\sum_{i = 1}^n 2^{t_i}},
    \\
    \sigma^2 &= \frac{\big(\sum_{i = 1}^n 2^{t_i}\big)
      \left[\sum_{i = 1}^n  2^{t_i}
        \big(\log \beta(t_i) - \mu\big)^2\right]}
    {\big(\sum_{i = 1}^n 2^{t_i}\big)^2
      - \sum_{i = 1}^n 2^{2 t_i}}.
  \end{split}
\end{equation}
A half-life of 1\;y means that the estimate of the transmission rate
from year $i$ is weighted half as much as the estimate from year
$i + 1$, a quarter as much as the estimate from year $i + 2$, and so
on.  This approach captured the recent country-level transmission
behavior while still allowing for the use of less-recent estimates to
smooth short-term variations (\autoref{transmission_rate}).

@@

Projections were generated for each country through model simulations
using 1000 random samples from published parameter distributions
(\autoref{model_param}) and the country-specific estimated
distribution of the transmission rate (\autoref{model_fitting}). The
parameters were sampled using Latin hypercube sampling
\cite{blower1994}.  We also used the variance reduction technique of
running different interventions with the same parameter samples, i.e.,
one set of sample parameter values is drawn, status quo and
intervention scenarios are all run with those sample values and
compared, and this process is repeated for each of the 1000 sample
sets of parameters \cite{shechter2006}.

The model outcomes were summarized using the median, 1st and 3rd
quartiles (i.e., 25th and 75th percentiles), and 5th and 95th
percentiles.  We calculated partial rank correlation coefficients
(PRCC) \cite{blower1994} to measure the independent effect of each
non-vaccine parameter on model projections of global new HIV
infections from 2015 to 2035 (\autoref{PRCCs}).

>>>>

\usepackage[font={sf}, labelfont={bf},
            labelsep=space, singlelinecheck=off]{caption}
\usepackage[font={sf}]{floatrow}
\usepackage{tabularx}

\usepackage[blocks]{authblk}
\renewcommand{\Authsep}{, }
\renewcommand{\Authand}{ \& }
\renewcommand{\Authands}{, }
\renewcommand{\Affilfont}{\small}

@@

\usepackage[style=nature]{biblatex}
\addbibresource{supplementary_information.bib}

\usepackage{xr}
\externaldocument{extended_data_fig_1}
\externaldocument{extended_data_fig_2}

@@

\renewcommand{\thesection}{S\arabic{section}}
\renewcommand{\figurename}{Supplementary Fig.}
\renewcommand{\figureautorefname}{\figurename}
\renewcommand{\tablename}{Supplementary Table}
\renewcommand{\tableautorefname}{\tablename}

\title{Supplementary Methods for\\
  \emph{Effectiveness of UNAIDS targets and HIV vaccination across 127
    countries}}

\author{Jan Medlock}
\affil{Department of Biomedical Sciences, Oregon State University, 106
  Dryden Hall, Corvallis, OR, 97331-4801, USA,
  \href{mailto:jan.medlock@oregonstate.edu}{\texttt{jan.medlock@oregonstate.edu}}}
\author{Abhishek Pandey} \author{Alyssa S.~Parpia} \author{Amber Tang}
\author{Laura A.~Skrip} \author{Alison P.~Galvani}
\affil{Center for Infectious Disease Modeling and Analysis, Yale
  School of Public Health, 135 College Street, New Haven, USA}

@@

The model simulation and analysis tools, written in Python, are
publicly available \autocite{medlock2016-git}.

@@

We developed a country-level continuous-time, compartmental model that
divides people aged $15$\;y and older into non-overlapping HIV states.
We chose to consider only these ages because they are the vast
majority of global PLHIV \autocite{UNICEF}, 93\% in 2014; because
population-level HIV estimates and demographic data are often
stratified into ages $15$--$49$; and to capture the continuing burden
of infections in people ages $50$ and older.

The background mortality rate ($\mu$) in each of the
127 countries was collected from published estimates
\autocite{World_Development_Indicators2013-ee}.  The recruitment rate
($\kappa$) into the age $15\;\text{y}$ and older group was set by the
difference between the country's population growth rate
\autocite{WorldBankpg} and its background mortality rate
\autocite{World_Development_Indicators2013-ee}, as would be true if the
population was at its stable age structure.

Our model has 8 HIV states (\autoref{model_diag}): susceptible to HIV
infection ($S$ is the number of people susceptible), vaccinated
against HIV ($R$), acute HIV infection ($A$), undiagnosed HIV
infection ($U$), diagnosed but untreated HIV infection ($D$), treated
without viral suppression ($T$), viral suppression ($V$), and having
AIDS ($W$).  Viral suppression is defined as a viral load of less than
1000 viral RNA copies per mL of blood.  Viral suppression not only
extends survival but also dramatically reduces transmission
\autocite{vernazza2000, ioannidis2001, attia2009, May2014-gp}.
Transitions between model states are governed by a series of
differential equations parametrized using values found in other
studies, along with transmission rates derived from UNAIDS estimates
of incidence and prevalence (\autoref{model_param}).  The model is
deterministic, but uncertainty in the parameters is treated by running
the model 1000 times with samples from the parameter distributions
(\autoref{uncertainty}) and summarizing the model outcomes using the
median, 1st and 3rd quartiles (i.e., 25th and 75th percentiles), and
5th and 95th percentiles.  Each country was modeled separately
(\hyperref[effectiveness_Afghanistan]{Extended Data
  Figs.~\ref*{effectiveness_Afghanistan}}--\ref{effectiveness_Zimbabwe})
and the country simulations were aggregated to calculate global and
regional outcomes
(\hyperref[effectiveness_Global]{Extended Data
  Figs.~\ref*{effectiveness_Global}}--\ref{effectiveness_Western_Europe}).

@@

\begin{table}
  \begin{center}
    \begin{tabularx}{\textwidth}{lXlll}
      \hline
      & Definition & Value & Reference \\
      \hline
      $\mu$ & Background death rate
      & Country specific, Supplementary Table 2
      & \autocite{World_Development_Indicators2013-ee} \\
      $\kappa$ & Recruitment rate
      & Country specific, Supplementary Table 2
      & \autocite{World_Development_Indicators2013-ee, WorldBankpg} \\
      $\delta$	& Rate of leaving acute infection
      & $\Triangular$(2, 4.14, 9.6)\;y$^{-1}$
      & \autocite{Hollingsworth2008-iy} \\
      $\sigma$	& Rate of developing AIDS without viral suppression
      & 0.1064\;y$^{-1}$ & \autocite{Morgan2002-cq} \\
      $\theta$ & Rate of developing AIDS with viral suppression
      & Eq.~(\ref{theta}) & --- \\
      $\gamma$ & Rate of viral suppression
      & $\Uniform$(0.5, 1.5)\;y$^{-1}$
      & \autocite{Currie2009-yz} \\
      $\nu$ & Death rate from AIDS & 0.5\;y$^{-1}$
      & \autocite{Morgan2002-cq} \\
      $\rho$ & Vaccine efficacy & 50\%\,[30\%, 70\%] & --- \\
      $\omega$	& Reduction in lifetime with viral suppression
      & $\Uniform$(5, 8)\;y
      & \autocite{Samji2013-kf, Unaids2014-ue} \\
      $\tau_{A}$ & Transmissibility during acute phase
      & $\Triangular$(0.0039, 0.0082, 0.0150)
      & \autocite{Wawer2005-us, Skarbinski2015-ni}\\
      $\tau_{U}$ & Transmissibility after acute phase
      & $\Triangular$(0.00077, 0.0014, 0.00251)
      & \autocite{Hughes2012-so} \\
      $\varepsilon$ & Relative transmissibility with
      viral suppression & $\BetaPERT$(0.08, 0.002, 0.57)
      & \autocite{Donnell2010-xo, attia_2009, wilson_2012, jia_2013,
        rodger_2016} \\
      $n$ & Coital acts per year & $\Uniform$(96, 108)
      & \autocite{Wawer2005-us, Abdool_Karim2010-cm}\\
      $\alpha$ & Diagnosis rate & See eq.~(\ref{diagnosis_rate}) & --- \\
      $\phi$ & Treatment rate & See eq.~(\ref{treatment_rate}) & --- \\
      $\psi$ & Rate of relapse to untreated & See eq.~(\ref{relapse_rate})
      & --- \\
      $\zeta$ & Vaccination rate & See eq.~(\ref{vaccination_rate}) & --- \\
      \hline
    \end{tabularx}
    \caption{Model parameters. $\Triangular$, $\Uniform$, and
      $\BetaPERT$ are sampling distributions: see
      \autoref{uncertainty} for a full description.  For vaccine
      efficacy, $50\%$ was the baseline, and $30\%$ and $70\%$ were
      also used in a scenario analysis.}
    \label{model_param}
  \end{center}
\end{table}

People susceptible to HIV ($S$) progress to acute infection ($A$)
according to the force of infection ($\lambda$,
eq.~\eqref{force_of_infection}), which depends on the HIV transmission
rate, estimated from the country-specific incidence and prevalence
data (\autoref{model_fitting}), adjusted for the current number of
acutely infected, untreated or ineffectively treated, and viral
suppression PLHIV.  Acute infection is characterized by high viral
loads and lasts an average of 2.90 [95\% confidence interval 1.23,
6.00]\;months \autocite{Hollingsworth2008-iy}.  (We converted these
durations to rates like $1 / 2.90\;\text{month$^{-1}$} \approx
4.14\;\text{y$^{-1}$}$ and set the parameter using random samples from
$\Triangular(2, 4.14, 9.6)\;\text{y$^{-1}$}$ to capture the
uncertainty in its value.  See \autoref{uncertainty} for the
definition of $\Triangular(a, b, c)$ and other distributions.)  We
assumed that people remain undiagnosed during their acute-infection
phase. After acute infection, people move into the chronic undiagnosed
class ($U$).  People in the undiagnosed class will be diagnosed at the
rate $\alpha$, which is determined by the country's current diagnosis
level and its diagnosis target (\autoref{targets}).  Diagnosed people
($D$) transition to the treated compartment ($T$) upon starting ART at
the rate $\phi$, dependent on the country's current treatment level
and its treatment target (\autoref{targets}).  After some time,
treated people then transition to the viral-suppression class ($V$),
at the rate $\gamma$.  We took $\gamma$ as a random sample from
$\Uniform(0.5, 1.5)\;\text{y$^{-1}$}$, that is, viral suppression
occurs after between 8 months and 2 years on treatment
\autocite{Currie2009-yz}.  Disengagement from treatment moves people from
treated ($T$) and viral suppression ($V$) at the rate $\psi$,
dependent on the current level of viral suppression and the country's
target (\autoref{targets}), back to the diagnosed class ($D$).

PLHIV without viral suppression (compartments $U$, $D$, and $T$)
develop AIDS (compartment $W$) after an average duration of 10.4 years
\autocite{Morgan2002-cq}, i.e., with transition rate
$\sigma = 0.1064\;\text{y$^{-1}$}$.  People who have achieved viral
suppression ($V$) develop AIDS at rate $\theta$, slower than PLHIV
without viral suppression.  People with viral suppression are
estimated to live $5$--$8$\;y shorter lives than the non-HIV infected
\autocite{Samji2013-kf, Unaids2014-ue}, which we quantified by
randomly sampling the parameter $\omega$ from
$\Uniform(5, 8)\;\text{y}$.  Assuming no relapse to untreated, the
duration of viral suppression is
\begin{equation}
  \frac{1}{\theta + \mu} = \frac{1}{\mu} - \omega - \frac{1}{\nu}.
\end{equation}
The term to the left of the equals sign is the time until leaving
viral suppression by either progression to AIDS (at rate $\theta$) or
background mortality (at rate $\mu$).  The term to the right of the
equals sign is the lifespan of the non-HIV infected, minus the
reduction in life due to having HIV with viral suppression, minus the
duration of the AIDS stage.  Solving for $\theta$ gives the rate of
progression to AIDS with viral suppression as
\begin{equation}
  \label{theta}
  \theta = \frac{1}{1/\mu - \omega - 1/\nu} - \mu.
\end{equation}

@@

The relative transmission rates of acute, unsuppressed, and suppressed
infected people are
\begin{equation}
  \label{betas}
  \begin{split}
    \beta_A &= 1 - (1 - \tau_A)^n,
    \\
    \beta_U &= 1 - (1 - \tau_U)^n,
    \\
    \beta_V &= 1 - (1 - \varepsilon \tau_U)^n,
  \end{split}
\end{equation}
where the $\tau$'s are the relative transmissibilities of people with
acute ($A$), unsuppressed ($U$ and $T$), and suppressed ($V$)
infections, $n$ is the annual number of coital acts, and $\varepsilon$
is the relative transmissibility with viral suppression.  For
transmissibility during the acute phase, we sampled $\tau_A$ from
$\Triangular$(0.0039, 0.0082, 0.0150)
\autocite{Wawer2005-us, Skarbinski2015-ni}, whereas, for transmissibility
for unsuppressed people, we sampled $\tau_U$ from
$\Triangular(0.00077, 0.0014, 0.00251)$ \autocite{Hughes2012-so}.  The
transmissibility of people with viral suppression relative to those
without suppression has been estimated at 0.08 (95\% confidence
interval 0.002, 0.57) \autocite{Donnell2010-xo}: due to the extremely long
left tail, we chose to sample the relative transmissibility
$\varepsilon$ from $\BetaPERT(0.08, 0.002, 0.57)$ rather than a
$\Triangular$ distribution.  (See \autoref{uncertainty} for a full
description of $\BetaPERT(a, b, c)$.)  The annual number of coital
acts $n$ was sampled from $\Uniform(96, 108)$
\autocite{Wawer2005-us, Abdool_Karim2010-cm}.  The country-specific
transmission rate $\eta$ was derived from UNAIDS estimates of each
country's longitudinal HIV prevalence and incidence
(\autoref{model_fitting}).

@@

The parametrized differential equations were solved numerically
\autocite[see][]{medlock2016-git} and four effectiveness outcomes were
computed from the solutions.
\begin{description}
\item[New infections] New infections from 2015 to time $t$ are given
  by $Y(t)$.

@@

The results from the country simulations were aggregated to the
regional and global levels by adding the numbers of people in each
compartment over time and then new infections, per-capita incidence,
PLHIV, and HIV-related deaths were calculated from the aggregates.
(See Supplementary Table 6 for the definitions of UNAIDS regions.)

@@

We modeled 3 different targets for the proportion diagnosed ($p_D$), the
proportion treated ($p_T$), and the proportion with viral suppression ($p_V$).
\begin{description}
\item[Status quo] The status quo target is for the proportions $p_D$,
  $p_T$, and $p_V$ to remain fixed at their initial levels
  (Supplementary Table 5) going forward. That is,
  \begin{equation}
    \label{status_quo_target}
    \begin{split}
      p_D^*(t) &= p_D(2015), \\
      p_T^*(t) &= p_T(2015), \\
      p_V^*(t) &= p_V(2015).
    \end{split}
  \end{equation}

@@

For 127 countries, we found sufficient published estimates and data to
parametrize our model.  Demographic data, including population growth
rate \autocite{WorldBankpg}, death rate
\autocite{World_Development_Indicators2013-ee}, and number of people
aged 15--49 years \autocite{The_World_Bank2016-fd} were obtained from
the World Bank (Supplementary Table 2). Longitudinal HIV prevalence
(Supplementary Table 3) and incidence (Supplementary Table 4)
estimates for ages 15--49, spanning from as early as 1990 to 2015,
were primarily obtained from the AIDSinfo database produced by UNAIDS
\autocite{Unaids2016-an}. Other sources included UNAIDS Country
Progress Reports \autocite{Unaids2016-am} published from 2012 to 2016
and AIDS Data Hub \autocite{AIDSdatahub-fg}, which compiles data from
UNAIDS, UNICEF, WHO, and the Asian Development Bank. These sources
were also used to inform the initial conditions for the model: the
number of people diagnosed with HIV, the number on ART, and the number
who have viral suppression or have been retained on treatment for at
least 12 months (Supplementary Table 5). Where estimates or data were
not available from these sources, alternative resources such as
peer-reviewed journal articles and country health ministry reports
were consulted.  See Supplementary Tables 2--5 for full information on
the sources used for each country.  These source tables are also
available in the public source-code repository
\autocite{medlock2016-git}.

@@

Using available historical country-level estimates of prevalence and
incidence (Supplementary Tables 3 and 4), and published estimates of
transmissibility
\autocite{Wawer2005-us,Donnell2010-xo,Hughes2012-so,Skarbinski2015-ni},
we derived an average country-specific transmission rate, which then
informs the acute, unsuppressed, and suppressed transmission rates.
This calibration method provided a recent estimate of country-specific
transmission, while using historical data to smooth out short-term
variations.

@@

To reflect the uncertainty in $\beta(t)$, we assumed that it follows a
lognormal distribution (\autoref{uncertainty}). For the parameters
$\mu$ and $\sigma^2$ of the lognormal distribution we used the
exponentially weighted mean and variance \autocite{holt2004} over time
(with half-life 1\;y) of $\log \beta(t)$ for the year 2015.  That is,
\begin{equation}
  \label{lognormal_params}
  \begin{split}
    \mu &= \frac{\sum_{i = 1}^n  2^{t_i}
      \log \beta(t_i)}
    {\sum_{i = 1}^n 2^{t_i}},
    \\
    \sigma^2 &= \frac{\big(\sum_{i = 1}^n 2^{t_i}\big)
      \left[\sum_{i = 1}^n  2^{t_i}
        \big(\log \beta(t_i) - \mu\big)^2\right]}
    {\big(\sum_{i = 1}^n 2^{t_i}\big)^2
      - \sum_{i = 1}^n 2^{2 t_i}}.
  \end{split}
\end{equation}
A half-life of 1\;y means that the estimate of the transmission rate
from year $i$ is weighted half as much as the estimate from year
$i + 1$, a quarter as much as the estimate from year $i + 2$, and so
on.  This approach captured the recent country-level transmission
behavior while still allowing for the use of less-recent estimates to
smooth short-term variations (\autoref{transmission_rate}).

@@

Projections were generated for each country through model simulations
using 1000 random samples from published parameter distributions
(\autoref{model_param}) and the country-specific estimated
distribution of the transmission rate (\autoref{model_fitting}). The
parameters were sampled using Latin hypercube sampling
\autocite{blower1994}.  We also used the variance reduction technique of
running different interventions with the same parameter samples, i.e.,
one set of sample parameter values is drawn, status quo and
intervention scenarios are all run with those sample values and
compared, and this process is repeated for each of the 1000 sample
sets of parameters \autocite{shechter2006}.

The model outcomes were summarized using the median, 1st and 3rd
quartiles (i.e., 25th and 75th percentiles), and 5th and 95th
percentiles.  We calculated partial rank correlation coefficients
(PRCC) \autocite{blower1994} to measure the independent effect of
each non-vaccine parameter on model projections of global new HIV
infections from 2015 to 2035
(\hyperref[PRCCs]{Extended Data Fig.~\ref*{PRCCs}}).

Given that vaccine parameters have not yet been quantified, we
restricted the number of unknowns by keeping vaccine efficacy constant
over time.  This assumption is based on proposed vaccine regimens that
include boosters for sustaining immunogenicity.  We chose plausible
baseline values for the included vaccine parameters (efficacy 50\%,
ultimate coverage 70\%, speed of rollout 25\% per year, year vaccine
is first available 2020) and examined uncertainty by simulating 6
additional vaccine scenarios, varying efficacy (30\%, 70\%), ultimate
coverage (50\%, 90\%), speed of rollout (10\% coverage per year), and
first year of availability (2025; Fig.~4). The vaccine scenario
analysis was done using the modes from the parameter distributions,
not samples from these distributions, to reduce the computation time.
